Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-07-08 main.tex
Modifs P.

% symbols
\usepackage{amsmath} 
\allowdisplaybreaks
\usepackage{amssymb} 
\usepackage{amsbsy}
\usepackage{bbold}
\usepackage{latexsym}
%\usepackage{amsfonts}
\usepackage{stmaryrd}
%\usepackage{mathabx}
%\usepackage{MnSymbol}
\usepackage{harmony} % simple music fonts
\usepackage{mathtools} % for arrows
%\usepackage{mathptmx}

@@

\begin{definition}
\label{def:transducer} \label{def:SWT}
A \emph{symbolic-weighted transducer} (\SWT)
over $\Sigma$, $\Delta$, $\Semiring$ and $\bar\Phi$
%the input and output alphabets~$\Sigma$ and $\Delta$ with label theory $\bar\Phi$, and the semiring $\Semiring$ 
is a tuple
$T = \< Q, \init, \bar{\wei}, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$   %\Phi_{\Sigma, \Delta}
(respectively $\mathsf{out} : Q \to \Semiring$)  %\Phi_{\Sigma, \Delta}
are functions defining the weight for entering
(respectively leaving) computation in a state,  
and $\bar{\wei}$ is a triplet of transition functions 
$\wei_{10}: Q \times Q \to \Phi_{\Sigma}$,
$\wei_{01}: Q \times Q \to \Phi_{\Delta}$, and 
$\wei_{11}: Q \times Q \to \Phi_{\Sigma, \Delta}$.
\end{definition}
%
\noindent
We call \emph{number of transitions} of $T$ the number of pairs of states 
$q, q' \in Q$ such that $\wei_{10}$ or $\wei_{01}$ or $\wei_{11}$
is not the constant $\zero$.
%
\noindent
For convenience, we shall sometimes present transitions 
as functions of  
\philippe{I missed sth: what is this $\varepsilon$? Intuitively clear but not defined?}
$Q \times (\Sigma \cup \{ \varepsilon \}) \times (\Delta \cup \{ \varepsilon \}) \times Q \to \Semiring$,
overloading the function names,
such that, for all $q, q' \in Q$, $a \in \Sigma$,  $b \in \Delta$,

@@

%
We recall that, by convention (Section~\ref{sec:semiring}), 
an empty sum with $\bigoplus$ is equal to~$\zero$. 
%
\philippe{OK tout ça se lit bien :-)}
Intuitively, using a transition $\wei_{ij}(q, a, b, q')$ means for $T$:
when reading respectively $a$ and $b$ at the current positions in the input and output words, 
increment the current position in the input word if and only if $i = 1$, 
and in the output word iff $j = 1$, %(otherwise, do not change it),
and change state from $q$ to $q'$.
When $a = \varepsilon$ (resp. $b = \varepsilon$), the current symbol 
in the input (resp. output) is not read.
%
%In contrast with the models of weighted transducers over finite alphabets~\cite{Mohri03ijfcs},
%the input and output symbols at current positions are always read by transitions, 
%even when they do not change the reading position the head's position.
%This is an important feature in the case of an infinite alphabet in 
%order to compare input and output symbols.
%which cannot be stored in the finite memory of the transducer.
%
Since~$\zero$ is absorbing for~$\otimes$ in~$\Semiring$,
one term $\wei_{ij}(q, a, b, q'')$ equal to $\zero$ in the above expression 
will be ignored in the sum, meaning that there is no possible transition
from state $q$ into state $q'$ while reading $a$ and $b$.
This is analogous to the case of a transition's guard not satisfied by $\<a, b>$ for 
symbolic transducers.

@@

\philippe{Je crois qu'il faudrait numéroter les exemples indépendamment des définitions. Cet exemple est le premier
qui donne des détails sur l'application visée. Il arrive peut-être un peu tard et est long. On pourrait introduire
la motivation dans l'intro, et développer des petits exemples au fur et à mesure.}
\begin{example}
In Common Western Music Notation~\cite{Gould11Notation}, 
several symbols may be used to represent one single sounding event. 
\lydia{unique $\to$ similar}
\florent{similar $\to$ single}
For instance, several notes can be combined with a tie, 
like in $\musQuarter\!\!\!\mathrel{\raisebox{-1.5mm}{$\smile$}}\!\!\!\musEighth$,  
and one note can be augmented by half its duration with a dot like in~\musQuarterDotted{}.
\lydia{modif.}
These notations are perceived equivalent when played, as their duration is equal, yet the notation is different.
We thus want to be able to compare a music score with music played by a performer. 
\lydia{changed end}
%
\noindent 
We propose a small weighted transducer model that calculates the distance bewteen an input sequence of sounding 
events (music "performance") 
to an output sequence of written events (music "score").
% simple pointwise distance between two sequences of timestamped events **
%
Let us consider the tropical (\emph{min-plus}) semiring~$\Semiring$ 
of Figure~\ref{fig:semirings} and 
let $\Sigma = \mathbb{R}_+$ be an input alphabet of event dates
and $\Delta = \{ \mathsf{e}, \mathsf{-} \} \times \mathbb{R}_+$ 
be an output alphabet of symbols with timestamps. 
A symbol $\< \mathsf{e}, d > \in \Delta$ represents an event starting at date $d$, 
and $\< \mathsf{-}, d >$ is a continuation of the previous event.
%More precisely, we let 
%$\Sigma = \{ a, - \} \times \R_+$ and 
%This example of $\Delta$ is motivated by the case of music notation, 
%where several notated events (notes) can be tied together, 
%with a \emph{tie} or a \emph{dot}
%meaning that they will be played as a unique sounding event.
%The timestamp of $a \in \Sigma$, denoted by $\mathsf{t}(a)$, is expressed as a rational number.

@@

%$\init(q_0, d, b) = \final(q_0, d, b) = \one$, and 
%$\init(q_1, d, b) = \final(q_1, d, b) = \zero$, 
%for all $d \in \Sigma$ and $b \in \Delta$.
\endex
\end{example}

@@

%\subsection{Properties}
\noindent
The following property will be useful to the approach on 
symbolic weighted parsing presented in Section~\ref{sec:parsing}.

@@

%\subsection{Definition}
\noindent
\label{sec:SWVPA-def}
Let $\Omega$ be a countable alphabet 
%finite (large) or infinite,
that we assume partitioned into three 
subsets~$\Omegai$, $\Omegac$, $\Omegar$,
whose elements are respectively called 
\emph{internal}, \emph{call} and \emph{return} symbols.
\philippe{Là je crois qu'il faudrait expliquer ces Omega, je commence à fatiguer et je suis un
peu largué par toutes ces définitions. J'intuite qu'il s'agit des symboles, parenthèses ouvrantes
et fermantes? Pourquoi il faut un alphabet pour les parenthèses?}
% names are  coined by application to functional program verification
% \begin{itemize}
% \item a set $\Omegai$ of \emph{internal symbols} denoted $a$,
% \item a set $\Omegac$ of \emph{call symbols} denoted $\call{a}$,
% \item a set $\Omegar$ of \emph{return symbols} denoted $\return{a}$.
% \end{itemize}
Let~$\< \Semiring, \oplus, \zero, \otimes, \one>$ be a commutative and complete semiring and let  
$\bar\Phi = \< \Phii, \Phic, \Phir, \Phici, \Phicc, \Phicr>$ 
be a label theory over $\Semiring$
%In order to simplify notations, %and following the definition of Section~\ref{section:transducer}, 
%we shall write respectively 
where $\Phii$, $\Phic$, $\Phir$ and~$\Phicx$ (with $\mathsf{x} \in \{ \mathsf{i}, \mathsf{c}, \mathsf{r}\}$) 
stand respectively 
for~$\Phi_\Omegai$, $\Phi_\Omegac$, $\Phi_\Omegar$ and~$\Phi_{\Omegac, \Omega_\mathsf{x}}$.
%
%Moreover, we extend this theory with a set $\Phii$ 
%of unary functions in $\Omegai \to \Semiring$,
%closed under $\oplus$ and $\otimes$.

\begin{definition}
A \emph{Symbolic Weighted Visibly Pushdown Automata} (\SWVPA) 
over  $\Omega = \Omegai \uplus \Omegac \uplus \Omegar$, $\Semiring$ and $\bar\Phi$ 
is a tuple $A = \< Q, P, \init, \bar\wei, \final >$,
where $Q$ is a finite set of states, 
\philippe{Est-ce que tout le monde sait ce qu'est un pushdown automata? Je suppose que c'est
lié à la pile.}
$P$ is a finite set of stack symbols, 
$\mathsf{in} : Q \to \Semiring$ 
(respectively $\mathsf{out} : Q \to \Semiring$)
are functions defining the weight for entering
(respectively leaving) a state, 
and $\bar\wei$ is a sextuplet composed of the transition functions :
$\weii : Q \times P \times Q \to \Phici$,  
$\weiei : Q \times Q \to \Phii$,  
$\weic : Q \times P \times Q \times P \to \Phicc$,  
$\weiec : Q \times P \times Q \to \Phic$,  
$\weir : Q \times P \times Q \to \Phicr$,  
$\weier : Q \times Q \to \Phir$.
%and 
%$\weiex : Q \times Q \to \Phix$ 
%with $\mathsf{x} \in \{ \mathsf{i}, \mathsf{c}, \mathsf{r}\}$.
\end{definition}
%
Similarly as in Section~\ref{section:transducer}, 
we extend the above transition functions as follows
for all $q, q' \in Q$, $p \in P$, 
$a \in \Omegai$, 
$\call{c} \in \Omegac$, 
$\return{r} \in \Omegar$, 
overloading their names: % for simplicity:
\[
\begin{array}{lll}
\weii: Q \times \Omegac \times P \times \Omegai \times Q \to \Semiring & 
\weii(q, c, p, a, q') = \eta_\mathsf{ci}(c, a) & 
\mathrm{where~} \eta_\mathsf{ci} = \weii(q, p, q'),\\
%
\weiei: Q \times \Omegai \times Q \to \Semiring & 
\weiei(q, a, q') = \phi_\mathsf{i}(a) &
\mathrm{where~} \phi_\mathsf{i} = \weiei(q, q').\\
%
\weic: Q \times \Omegac \times P \times  \Omegac \times P \times Q \to \Semiring & 
\weic(q, c, p, \call{c'}, p', q') = \eta_\mathsf{cc}(c, \call{c'}) & 
\mathrm{where~} \eta_\mathsf{cc} = \weic(q, p, p', q'),\\
%
\weiec: Q \times \Omegac \times P \times Q \to \Semiring & 
\weiec(q, {\call{c}}, p, q') = \phi_\mathsf{c}({\call{c}}) &
\mathrm{where~} \phi_\mathsf{c} = \weiec(q, p, q').\\
%
\weir: Q \times \Omegac \times P \times \Omegar \times Q \to \Semiring & 
\weir(q, {\call{c}},  p, {\return{r}}, q') = \eta_\mathsf{cr}({\call{c}},  {\return{r}}) & 
\mathrm{where~} \eta_\mathsf{cr} = \weir(q, p, q'),\\
%
\weier: Q \times \Omegar \times Q \to \Semiring & 
\weier(q, {\return{r}}, q') = \phi_\mathsf{r}({\return{r}}) &
\mathrm{where~} \phi_\mathsf{r} = \weier(q, q').\\
\end{array}      
\]

>>>>

% symbols
\usepackage{amsmath}
\allowdisplaybreaks
\usepackage{amssymb}
\usepackage{amsbsy}
\usepackage{bbold}
\usepackage{latexsym}
%\usepackage{amsfonts}
\usepackage{stmaryrd}
%\usepackage{mathabx}
%\usepackage{MnSymbol}
\usepackage{harmony} % simple music fonts
\usepackage{mathtools} % for arrows
%\usepackage{mathptmx}
\usepackage{comment}

@@

\begin{definition}
\label{def:transducer} \label{def:SWT}
A \emph{symbolic-weighted transducer} (\SWT)
over $\Sigma$, $\Delta$, $\Semiring$ and $\bar\Phi$
%the input and output alphabets~$\Sigma$ and $\Delta$ with label theory $\bar\Phi$, and the semiring $\Semiring$
is a tuple
$T = \< Q, \init, \bar{\wei}, \final >$,
where $Q$ is a finite set of states,
$\mathsf{in} : Q \to \Semiring$   %\Phi_{\Sigma, \Delta}
(respectively $\mathsf{out} : Q \to \Semiring$)  %\Phi_{\Sigma, \Delta}
are functions defining the weight for entering
(respectively leaving) computation in a state,
and $\bar{\wei}$ is a triplet of transition functions
$\wei_{10}: Q \times Q \to \Phi_{\Sigma}$,
$\wei_{01}: Q \times Q \to \Phi_{\Delta}$, and
$\wei_{11}: Q \times Q \to \Phi_{\Sigma, \Delta}$.
\end{definition}
%
\noindent
We call \emph{number of transitions} of $T$ the number of pairs of states
$q, q' \in Q$ such that $\wei_{10}$ or $\wei_{01}$ or $\wei_{11}$
is not the constant $\zero$.
%
\noindent
For convenience, we shall sometimes present transitions
as functions of
$Q \times (\Sigma \cup \{ \varepsilon \}) \times (\Delta \cup \{ \varepsilon \}) \times Q \to \Semiring$,
overloading the function names,
such that, for all $q, q' \in Q$, $a \in \Sigma$,  $b \in \Delta$,

@@

%
We recall that, by convention (Section~\ref{sec:semiring}),
an empty sum with $\bigoplus$ is equal to~$\zero$.
%
Intuitively, using a transition $\wei_{ij}(q, a, b, q')$ means for $T$:
when reading respectively $a$ and $b$ at the current positions in the input and output words,
increment the current position in the input word if and only if $i = 1$,
and in the output word iff $j = 1$, %(otherwise, do not change it),
and change state from $q$ to $q'$.
When $a = \varepsilon$ (resp. $b = \varepsilon$), the current symbol
in the input (resp. output) is not read.
%
%In contrast with the models of weighted transducers over finite alphabets~\cite{Mohri03ijfcs},
%the input and output symbols at current positions are always read by transitions,
%even when they do not change the reading position the head's position.
%This is an important feature in the case of an infinite alphabet in
%order to compare input and output symbols.
%which cannot be stored in the finite memory of the transducer.
%
Since~$\zero$ is absorbing for~$\otimes$ in~$\Semiring$,
one term $\wei_{ij}(q, a, b, q'')$ equal to $\zero$ in the above expression
will be ignored in the sum, meaning that there is no possible transition
from state $q$ into state $q'$ while reading $a$ and $b$.
This is analogous to the case of a transition's guard not satisfied by $\<a, b>$ for
symbolic transducers.

@@

\begin{example}
Let us develop the example of comparison between music played by a performer,
represented as a sequence $s \in \Sigma^*$ of events in the MIDI alphabet $\Sigma$,
and  a music score represented as a sequence $t \in \Delta^*$
in the CWMN alphabet $\Delta$. We build a small weighted transducer model
 with two states $q_0$ and $q_1$ that
calculates the distance between $s$ and $t$.

If one performed event $s_i$  corresponds
to one notated event $t_1$ (for instance MIDI code 61 and pitch A4),
the weight value computed by the \SWT is the time distance between both,
as in Example~\ref{distance-time}, and is modeled by
transitions $\wei_{11}$ below.
%
If we meet the music notation symbol '-' that
represents continuation (such as instance in \emph{ties}
$\musQuarter\!\!\!\mathrel{\raisebox{-1.5mm}{$\smile$}}\!\!\!\musEighth$,
or \emph{dots} \musQuarterDotted{}), it is  skipped with no cost (transitions $\wei_{01}$ or weight $\one$).
%transitions
\[
\begin{array}{rclcrcl}
\wei_{11}(q_0, d, \< \mathsf{e}, d'>, q_0) & = & |d' - d| & \quad &
\wei_{11}(q_1, d, \< \mathsf{e}, d'>, q_0) & = & |d' - d|\\
\wei_{01}(q_0, \varepsilon, \< \mathsf{-}, d'>, q_0) & = & \one & &
\wei_{01}(q_1, \varepsilon, \< \mathsf{-}, d'>, q_0) & = & \one\\
\wei_{10}(q_0, d, \varepsilon, q_1) & = & \alpha & & %\multicolumn{3}{l}{\mathrm{for~all~} b \in \Delta}
\end{array}
\]
%
We also must be able to take performing errors into account, while still being able to compare with the score,
since a performer could, for example, play an unwritten extra note.
\lydia{reformulated this sentence}
%
This is modelled by the transition $\wei_{10}$ with an arbitrary weight value $\alpha \in \Semiring$,
switching from state $q_0$ (normal) to $q_1$ (error).
\philippe{Comprends pas cette phrase}
The transitions in the second column below switch back to the normal state $q_0$.
% the metric computed by the \SWT is the smallest sum of point wise distances
% between dates of input and output events.
At last, we let $q_0$ be the only initial and final state, with
$\init(q_0) = \final(q_0) = \one$, and
$\init(q_1) = \final(q_1) = \zero$.

\lydia{ccl to the ex}
That way, an \SWT is capable of evaluating the differences between a score and a performance,
all the while ensuring that performance errors are plausible.
%$\init(q_0, d, b) = \final(q_0, d, b) = \one$, and
%$\init(q_1, d, b) = \final(q_1, d, b) = \zero$,
%for all $d \in \Sigma$ and $b \in \Delta$.
\endex
\end{example}

\begin{comment}
\begin{example}
In Common Western Music Notation~\cite{Gould11Notation},
several symbols may be used to represent one single sounding event.
\lydia{unique $\to$ similar}
\florent{similar $\to$ single}
For instance, several notes can be combined with a tie,
like in $\musQuarter\!\!\!\mathrel{\raisebox{-1.5mm}{$\smile$}}\!\!\!\musEighth$,
and one note can be augmented by half its duration with a dot like in~\musQuarterDotted{}.
\lydia{modif.}
These notations are perceived equivalent when played, as their duration is equal, yet the notation is different.
We thus want to be able to compare a music score with music played by a performer.
\lydia{changed end}
%
\noindent
We propose a small weighted transducer model with two states $q_0$ and $q_1$
that calculates the distance bewteen an input sequence of sounding
events (music "performance")
to an output sequence of written events (music "score").
% simple pointwise distance between two sequences of timestamped events **
%
Let us consider the tropical (\emph{min-plus}) semiring~$\Semiring$
of Figure~\ref{fig:semirings} and
let $\Sigma = \mathbb{R}_+$ be an input alphabet of event dates
and $\Delta = \{ \mathsf{e}, \mathsf{-} \} \times \mathbb{R}_+$
be an output alphabet of symbols with timestamps.
A symbol $\< \mathsf{e}, d > \in \Delta$ represents an event starting at date $d$,
and $\< \mathsf{-}, d >$ is a continuation of the previous event.
%More precisely, we let
%$\Sigma = \{ a, - \} \times \R_+$ and
%This example of $\Delta$ is motivated by the case of music notation,
%where several notated events (notes) can be tied together,
%with a \emph{tie} or a \emph{dot}
%meaning that they will be played as a unique sounding event.
%The timestamp of $a \in \Sigma$, denoted by $\mathsf{t}(a)$, is expressed as a rational number.

@@

%$\init(q_0, d, b) = \final(q_0, d, b) = \one$, and
%$\init(q_1, d, b) = \final(q_1, d, b) = \zero$,
%for all $d \in \Sigma$ and $b \in \Delta$.
\endex
\end{example}
\end{comment}

@@

%\subsection{Properties}
\noindent
\philippe{Il me manque une explication:  on construit un automate
qui, étant donnée une partition $t$, renvoie la distance minimale
avec n'importe quelle performance (distance donnée par un transducer)?
Quel est le rôle de A(s)?
}
The following property will be useful to the approach on
symbolic weighted parsing presented in Section~\ref{sec:parsing}.

@@

%\subsection{Definition}
\noindent
\label{sec:SWVPA-def}
Let $\Omega$ be a countable alphabet
%finite (large) or infinite,
that we assume partitioned into three
subsets~$\Omegai$, $\Omegac$, $\Omegar$,
whose elements are respectively called
\emph{internal}, \emph{call} and \emph{return} symbols.
% names are  coined by application to functional program verification
% \begin{itemize}
% \item a set $\Omegai$ of \emph{internal symbols} denoted $a$,
% \item a set $\Omegac$ of \emph{call symbols} denoted $\call{a}$,
% \item a set $\Omegar$ of \emph{return symbols} denoted $\return{a}$.
% \end{itemize}
Let~$\< \Semiring, \oplus, \zero, \otimes, \one>$ be a commutative and complete semiring and let
$\bar\Phi = \< \Phii, \Phic, \Phir, \Phici, \Phicc, \Phicr>$
be a label theory over $\Semiring$
%In order to simplify notations, %and following the definition of Section~\ref{section:transducer},
%we shall write respectively
where $\Phii$, $\Phic$, $\Phir$ and~$\Phicx$ (with $\mathsf{x} \in \{ \mathsf{i}, \mathsf{c}, \mathsf{r}\}$)
stand respectively
for~$\Phi_\Omegai$, $\Phi_\Omegac$, $\Phi_\Omegar$ and~$\Phi_{\Omegac, \Omega_\mathsf{x}}$.
%
%Moreover, we extend this theory with a set $\Phii$
%of unary functions in $\Omegai \to \Semiring$,
%closed under $\oplus$ and $\otimes$.

\begin{definition}
A \emph{Symbolic Weighted Visibly Pushdown Automata} (\SWVPA)
over  $\Omega = \Omegai \uplus \Omegac \uplus \Omegar$, $\Semiring$ and $\bar\Phi$
is a tuple $A = \< Q, P, \init, \bar\wei, \final >$,
where $Q$ is a finite set of states,
lié à la pile.}
$P$ is a finite set of stack symbols,
$\mathsf{in} : Q \to \Semiring$
(respectively $\mathsf{out} : Q \to \Semiring$)
are functions defining the weight for entering
(respectively leaving) a state,
and $\bar\wei$ is a sextuplet composed of the transition functions :
$\weii : Q \times P \times Q \to \Phici$,
$\weiei : Q \times Q \to \Phii$,
$\weic : Q \times P \times Q \times P \to \Phicc$,
$\weiec : Q \times P \times Q \to \Phic$,
$\weir : Q \times P \times Q \to \Phicr$,
$\weier : Q \times Q \to \Phir$.
%and
%$\weiex : Q \times Q \to \Phix$
%with $\mathsf{x} \in \{ \mathsf{i}, \mathsf{c}, \mathsf{r}\}$.
\end{definition}
%
Similarly as in Section~\ref{section:transducer},
we extend the above transition functions as follows
for all $q, q' \in Q$, $p \in P$,
$a \in \Omegai$,
$\call{c} \in \Omegac$,
$\return{r} \in \Omegar$,
overloading their names: % for simplicity:
\[
\begin{array}{lll}
\weii: Q \times \Omegac \times P \times \Omegai \times Q \to \Semiring &
\weii(q, c, p, a, q') = \eta_\mathsf{ci}(c, a) &
\mathrm{where~} \eta_\mathsf{ci} = \weii(q, p, q'),\\
%
\weiei: Q \times \Omegai \times Q \to \Semiring &
\weiei(q, a, q') = \phi_\mathsf{i}(a) &
\mathrm{where~} \phi_\mathsf{i} = \weiei(q, q').\\
%
\weic: Q \times \Omegac \times P \times  \Omegac \times P \times Q \to \Semiring &
\weic(q, c, p, \call{c'}, p', q') = \eta_\mathsf{cc}(c, \call{c'}) &
\mathrm{where~} \eta_\mathsf{cc} = \weic(q, p, p', q'),\\
%
\weiec: Q \times \Omegac \times P \times Q \to \Semiring &
\weiec(q, {\call{c}}, p, q') = \phi_\mathsf{c}({\call{c}}) &
\mathrm{where~} \phi_\mathsf{c} = \weiec(q, p, q').\\
%
\weir: Q \times \Omegac \times P \times \Omegar \times Q \to \Semiring &
\weir(q, {\call{c}},  p, {\return{r}}, q') = \eta_\mathsf{cr}({\call{c}},  {\return{r}}) &
\mathrm{where~} \eta_\mathsf{cr} = \weir(q, p, q'),\\
%
\weier: Q \times \Omegar \times Q \to \Semiring &
\weier(q, {\return{r}}, q') = \phi_\mathsf{r}({\return{r}}) &
\mathrm{where~} \phi_\mathsf{r} = \weier(q, q').\\
\end{array}
\]
